Embracing the Laws of Physics: \\ Reversible Models of Computation

### 2017-11-24 survey/survey.tex
introduction that can lead into the paper as currently envisioned.

Many resolutions were offered for this conundrum (for a compilation,
see Maxwell's Demon by Leff and Rex~\cite{leff1990}), but none withstood careful
scrutiny until the establishment of the so called `Landauer's
Principle' in 1961 -- a principle whose experimental validation
happened recently in 2012 by Eric Lutz and others.

Maxwell's demon appears to violate the second law of thermodynamics by
having a tiny `intelligence' observing the movement of individual
particles of a gas and separating fast moving particles from slow
moving ones, thereby reducing the total entropy of the
system. Landauer's resolution of the demon relied on two ideas that
took root only a few decades earlier -- the formal notion of
computation (through the work of Turing and Church, 1936) and the
formal notion of information (through the work of Claude Shannon,
1948). Landauer reasoned that the computation done by the finite brain
of the demon, involves getting information about the movement of
molecules, acting on that information and then overwriting it to make
room for the next computation.  Landauer reasoned, and this is the
important bit, that the computation that is manipulating information
in the demon's brain \textit{must be thermodynamic work}, thereby
bringing the demon back into the fold of physics.

In its essence this is a strange and wonderful idea. Information,
physics and computation are inextricably linked and that ideas in each
field have implications for the other. (Cite Bennett and the various
`thermodynamics of computation' papers here.)

When the early models computation, the Turing machine, and the
$\lambda$-calculus, were developed there was no real reason to the
take the information content of computations into serious
consideration -- in fact, at that time there was no quantifiable
notion of information. These models followed in the footsteps of logic
where conventional logic gates such as AND and OR happily erase their
inputs.

@@

Now however, through the establishment of the Landauer's principle and
through recent investigations into computational models that are well
suited to quantum mechanics it has become increasingly evident that
our computation models need to be revisited. Information, not just
seems to have a physical manifestation, but also that it, like other
physical entities, is conserved.

Going back to the drawing board, can we have computation that closely
matches our underlying physics?  i.e. can we have computation that in
its essence preserves information? What is the missing principle that
we can use as a basis of a new model of computation to achieve this?
It turns out that we need to look at physics again for inspiration --
in physics all of the primitive rules, nature itself at its most
fundamental level, is reversible. Conventional computation is not. Can
we embrace this simple principle as the building block of a model of
computation? What would computation look like when viewed from this
vantage point? And, does it have applications?

The objective of this survey is to touch upon the many related areas
of research in this space.

%% Abstract models of computation: Turing machine, RAM, Von Neumann
%% architecture, $\lambda$-calculus, logic, discrete mathematics

%% \jc{why are logic and DM listed alongside MoC?}

>>>>

Many resolutions were offered for this conundrum (for a compilation, see
Maxwell's Demon by Leff and Rex~\cite{leff1990}), but none withstood careful
scrutiny until the establishment of `Landauer's Principle' in 1961 -- a
principle whose experimental validation
happened recently in 2012~\cite{lutz2012}.

Maxwell's demon appears to violate the second law of thermodynamics by
having a tiny `intelligence' observing the movement of individual
particles of a gas and separating fast moving particles from slow
moving ones, thereby reducing the total entropy of the
system. Landauer's resolution of the demon relied on two ideas that
took root only a few decades earlier -- the formal notion of
computation (through the work of Turing and Church, 1936) and the
formal notion of information (through the work of Claude Shannon,
1948). Landauer reasoned that the computation done by the finite brain
of the demon, involves getting information about the movement of
molecules, acting on that information and then overwriting it to make
room for the next computation.  Landauer reasoned, and this is the
important part, that the computation that is manipulating information
in the demon's brain \textit{must be thermodynamic work}, thereby
bringing the demon back into the fold of physics.

This is a strange and wonderful idea: Information,
physics and computation are inextricably linked. Furthermore, this
implies that ideas in each field have consequences for the other. 
(Cite Bennett and the various `thermodynamics of computation' papers here.)

When the early models computation, the Turing machine, and the
$\lambda$-calculus, were developed, there was no compelling reason to the
take the information content of computations into
consideration -- in fact, at that time there was no quantifiable
notion of information. These models followed in the footsteps of logic
where conventional gates such as AND and OR happily erase their
inputs. These models were already radical enough, by being purely
constructive, in an era where classical mathematics, with its
pervasive use of excluded middle and frequent use of the axiom of
choice, did not overly worry about effective representation.

@@

In our current age, both Laudauer's Princple and recent investigations
of computational models well-suited to quantum mechanics, push us to
revisit our current models of computation. Information, on top of
having a physical manifestation, appears to be \emph{conserved},
along with other quantities like energy, mass and momentum.

But what does it mean for a computation to preserve information?
What is the missing principle that will guide the creation of a new
model of computation that ``conserve information''?
It is nature to look again at physics for inspiration --
in our current understanding of physics all of the primitive rules
are reversible. Conventional computation is not. Can
we embrace this simple principle as the building block of a model of
computation? What would computation look like when viewed from this
vantage point? Can non-trivial computation be performed in a
reversible manner? And, if so, does it have applications?

Traditional models of computation are all unityped, generally
operating on unstructured sequences of symbols. But recent
investigations, in the intersection between mathematics and
computer science (particularly in Homotopy Type Theory~\cite{HoTT-book})
reveal that not only can mathematics be typed, but that doing so
reveals further structure -- in both mathematics and in computation.

And that brings us to the objective of this survey: explore the domain of
\emph{typed reversible computation}.
